---
title: "Reproducible Research Project 2"
author: Gabriela Ochoa
output: 
  html_document:
    keep_md: true
---

# **Introduction**
Storms and other severe weather events can cause both public health and economic problems for communities and municipalities. Many severe events can result in fatalities, injuries, and property damage, and preventing such outcomes to the extent possible is a key concern.

This project involves exploring the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. This database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage.

## Data
The data for this assignment come in the form of a comma-separated-value file compressed via the bzip2 algorithm to reduce its size. You can download the file from the course web site:

- [Storm Data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2)

The basic goal of this assignment is to explore the NOAA Storm Database and answer some basic questions about severe weather events. You must use the database to answer the questions below and show the code for your entire analysis. Your analysis can consist of tables, figures, or other summaries. You may use any R package you want to support your analysis.

## *Questions*

    1. Across the United States, which types of events (as indicated in the **EVTYPE** variable) are most harmful with respect to population health?  
    2. Across the United States, which types of events have the greatest economic consequences?


# **Data Processing**
```{r}
if (!file.exists("StormData.csv.bz2")) {
    fileURL <- 'https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2'
    download.file(fileURL, destfile='StormData.csv.bz2', method = 'curl')
}
noaaDF <- read.csv(bzfile('StormData.csv.bz2'),header=TRUE, stringsAsFactors = FALSE)
```
### Load Packages
```{r, echo=FALSE, results='hide', warning=FALSE, message=FALSE}
library(dplyr)
library(tidyr)
library(lubridate)
library(ggplot2)
library(gridExtra)

```

### exploring data
```{r, echo=FALSE, results='hide', warning=FALSE, message=FALSE}
summary(noaaDF)
str(noaaDF)
```

We can select the variables that will help us to analyze the damage based on EVTYPE (event type), FATALITIES, INJURIES, PROPDMG (property damage), PROPDMGEXP (property damage expense), CROPDMG (crop damage), and CROPDMGEXP (crop damage expense). 
```{r}
NOAA <- noaaDF[,c('EVTYPE','FATALITIES','INJURIES', 'PROPDMG', 'PROPDMGEXP', 'CROPDMG', 'CROPDMGEXP')]

## View headers in revised data set
head(NOAA)
```

```{r}
## Organize type of event ~ fatalities and store in object "fatalities" & same for "injuries"
fatalities <- aggregate(FATALITIES ~ EVTYPE, data=NOAA, sum)
injuries <- aggregate(INJURIES ~ EVTYPE, data = NOAA, sum)

## Sort fatalities
fatalities <- fatalities[order(-fatalities$FATALITIES), ][1:20, ]
fatalities$EVTYPE <- factor(fatalities$EVTYPE, levels = fatalities$EVTYPE)
head(fatalities)
```

```{r}
## Sort Injuries
injuries <- injuries[order(-injuries$INJURIES), ][1:20, ]
injuries$EVTYPE <- factor(injuries$EVTYPE, levels = injuries$EVTYPE)
head(injuries)

```
# **Results**  

## *Question 1*
**Across the United States, which types of events (as indicated in the *EVTYPE* variable) are most harmful with respect to population health?**
```{r}
p1 = ggplot(fatalities, aes(x = EVTYPE, y = FATALITIES, theme_set(theme_bw()))) + 
    geom_bar(stat = "identity", fill = "steelblue") + 
    theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 6)) + 
    xlab("Event Type") + ylab("Fatalities") + ggtitle("Fatalities by top 20 Weather Event Types") +
    theme(plot.title = element_text(size = 10))

p2 = ggplot(injuries, aes(x = EVTYPE, y = INJURIES, theme_set(theme_bw()))) + 
    geom_bar(stat = "identity", fill = "plum4") + 
    theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 6)) + 
    xlab("Event Type") + ylab("Injuries") + ggtitle("Injuries by top 20 Weather Event Types") +
    theme(plot.title = element_text(size = 10))

## Plot both side by side using gridExtra package
grid.arrange(p1, p2, ncol = 2, top = "Most Harmful Events with Respect to Population Health")
```

As show in the plot Tornado is the event type that has the highest level of Fatalities and Injuries.

### *Question 2*
**Across the United States, which types of events have the greatest economic consequences?** 

We proceed to calculate the costo of property and crop damages.

```{r}
## property
prop <- aggregate(NOAA$PROPDMG, by = list(NOAA$EVTYPE), "sum")
names(prop) <- c("Event", "Property")
prop_sorted <- prop[order(-prop$Property), ][1:20, ]
head(prop_sorted)
```
Crop

```{r}
crop <- aggregate(NOAA$CROPDMG, by = list(NOAA$EVTYPE), "sum")
names(crop) <- c("Event", "Crop")
crop_sorted <- crop[order(-crop$Crop), ][1:20, ]
head(crop_sorted)
```
### **Plot**


```{r}
par(mfrow = c(1, 2), mar = c(10, 4, 2, 2), las = 3, cex = 0.7, cex.main = 1.4, cex.lab = 1.2)

barplot(prop_sorted$Property, names.arg = prop_sorted$Event, col= "lightseagreen", main="Top Weather Events for Poperty Damage", ylab="Amout of Property Damage", ylim=c(0,3500000))

barplot(crop_sorted$Crop, names.arg = crop_sorted$Event, col= "hotpink", main="Top Weather Events for Crop Damage", ylab="Amout of Crop Damage", ylim=c(0,3500000))

```
```{r}
## Considering both, property and crop
TotalCost <- aggregate(NOAA$CROPDMG+NOAA$PROPDMG, by = list(NOAA$EVTYPE), "sum")
names(TotalCost) <- c("Event", "TotalCost")
total_sorted <- TotalCost[order(-TotalCost$TotalCost), ][1:20, ]
head(total_sorted)
```

```{r}
barplot(total_sorted$TotalCost, names.arg = total_sorted$Event, col = 'palevioletred4',
        main = 'Top Weather Events for total Damage ', ylab = 'Amount of total Damage', ylim = c(0, 3500000))
```

We can confirm that Tornadoes are the main cause for damage. 